Plant species traits and growth rates: meta-analysis shows correlations change with plant size as predicted

### 2015-08-07 MS.tex
Change in the meta-analysis

\item
The literature is inconsistent about empirical correlations between functional traits and plant growth rate (GR), casting doubt on the capacity of traits to predict growth.

@@

\section*{Introduction}\label{introduction}
Plant growth is an important aspect of ecosystem productivity
Plant species vary considerably in the growth rate they can achieve under favourable growing conditions (Grime and Hunt 1975; Shipley and Keddy 1988; Poorter and Retakes 1990; Lambers and Poorter 1992). This variation is important both because potential RGR is an important aspect of ecological performance, and because slow potential RGR is thought to be part of a nexus of adaptations to unfavourable growing conditions (e.g. Grime et al. 1988; Reich et al. 1992; Chapin et al. 1993; Aerts and van der Pijl 1993; van der Werf et al. 1993).

The phrase "plant ecological strategy" refers to how species face the challenge they all share:  to acquire sufficient water, nutrients and light for individual growth and to live long enough to replace themselves with offspring. An idea that species strategy differences might involve fast vs slow growth goes back at least to Grime and Hunt \citeyearpar{Grime:1975gr}, who measured a wide spread of relative growth rates for seedlings during early exponential growth under favourable conditions. Seedling potential relative growth rate (RGR, the change in a mass/height or diamater per time interval relative to initial size) has continued to be used as a major strategy indicator within the CSR (Competitor - Stress tolerator - Ruderal) scheme and its "Integrated Screening Program" (e.g. \citealt{grime1979plant, Grime:1997wm}). Meanwhile partitioning of seedling RGR into its components (see equation X in Appendix 1) showed that specific leaf area (SLA = 1/leaf mass per area LMA) was typically the strongest source of variation between species \citep{Poorter:1989tx}. From the 1990s a new style of "trait ecology" arose whereby measurable traits including SLA were used directly as strategy axes \citep{Westoby:2002ft}. An approach via measurable traits made broad comparisons possible across continents and latitudes and thousands of species. Traits have been seen as constructive approaches to understanding plant growth strategies \citep{Grime:1977kc,Chapin:1980gz}, community composition \citep{Lavorel:2002ff,Shipley:2006ie}, global vegetation dynamics \citep{Scheiter:2013ed} and ecosystem processes \citep{Lavorel:2002ff}.

Currently two main spectra of variation are widely cited as underpinning differences between species in vegetative growth rates. The `leaf economic spectrum' \citep{Wright:2004jb} and the 'wood economic spectrum' \citep{Chave:2009iy} reflect tissue construction costs (for specific leaf area and for wood volume respectively) trading off with tissue turnover rates or mortality risks. It has been widely expected that species with low tissue construction cost will typically have fast growth rates at least in favourable physical environments (e.g. \citealt{MullerLandau:2004dc,Wright:2004jb,Poorter:2008iu,Chave:2009iy,Larjavaara:2010bn,Iida:2012jb,Paine:2015df}). Yet a growing number of empirical studies indicate the correlation between traits and growth is not as consistent as that. Consider SLA as an illustration. Species with high SLA can deploy more canopy area for a given dry mass investment in leaves \citep{Poorter:1999wd, Reich:1992wm}. In addition, SLA is correlated with high photosynthetic rates and high nitrogen (N) or phosphorus concentrations in the leaf \citep{Wright:2004jb}. For seedling relative growth rates under favourable conditions, high SLA has repeatedly been found to be a strong predictor regardless of vegetation type or growth form \citep{Lambers:1992bj,Reich:1992wm,Grime:1997wm,Poorter:1999wd,Wright:1999ds}. Yet for adult plants, this correlation appears weaker or absent \citep{coomes_comparison_1998,Poorter:2008iu,Aiba:2009ft,Easdale:2009gv,Wright:2010tp}. The growing discrepancy between theoretical expectation and empirical results has led some to question whether traits are even useful for understanding growth \citep{Wright:2010tp, Paine:2015df}.

Meanwhile the possibility has been raised that trait-growth correlations may change systematically with size \citep{Falster:2011ii, Ruger:2012jv, Iida:2014ep, Iida:2014hq}.ADD REF  Recent theory suggests that the absence of a correlation between SLA and growth at larger sizes may be expected, when potential influences of traits on plant function are considered \citep{Falster:2011ii, falster:2013}. At larger sizes the low-cost deployment benefits of high SLA may be offset or even outweighed by the high costs of rapid leaf turnover (see Text Box 1\ref{sec:growth}).

@@

The Web of Science database (Thomson Reuters) was searched between March and June 2014 using a range of Boolean search terms in English. Search terms were as follows: (''NAME'' AND (''growth rate'' OR ''relative growth rate'')) AND (''tree*'' OR ''woody*''), ''NAME'' was one of trait names listed in \ref{TableA1}. According to this snapshot, 62\% of the literature for the 20 traits (or 82\% for the 5 traits for which we have clear predictions) is captured by searching only on SLA and WD.

@@

Google Scholar database was searched using the range of Boolean search terms for SLA and WD. Search terms were as follows: ((''SLA'' OR ''LMA'' OR ''specific leaf area'' OR ''leaf mass area'' OR ''wood density'' OR ''WD'' OR ''wood specific gravity'' OR ''stem density'' OR ''stem specific density'') AND (''growth rate'' OR ''relative growth rate'')) AND (tree* OR woody*). The keyword search in Google Scholar returns approx. 18900 papers. For comparison, with the same keyword search Web of Science returns 541 papers. Thus, in Google Scholar we displayed papers by relevance and examined the first 750 items.

Each study had to meet some basic criteria to be included in our analysis. We restricted our meta-analysis to documents reporting: i) the relationship between growth and trait as a correlation coefficient, ii) the sample size and the number of species used to perform the correlation, and iii) the size/stage of plants. Where available we also recorded the geographic coordinates, number of sites, experiment type (greenhouse, field experiment, forest, shade-house etc.), growth measure (relative or absolute growth rate measured using plant diameter, height or mass), vegetation type, life form (tree, woody or mixed with herbaceous species etc.), biological scale (interspecific, intraspecific or intragenus), and experimental treatments (e.g.~fertilization, shade, salt stress). Ultimately data was included from 112 different studies (starred references in bibliography). The articles laid aside because they did not meet our criteria are listed in Appendix \ref{app:supp_info_files} in Supplementary material.

The number of species used to perform the correlation, the experimental conditions, and the concordance between stage used to measure trait and growth are clear criteria for assessing the quality of the data for addressing our hypotheses. Therefore, we focused on correlation across species, and we stratified the data set into two subsets: one including exclusively studies under unstressed conditions ('ideal' dataset), and the other set including studies under all conditions ('complete' dataset). We excluded from the 'ideal' dataset studies where growth and traits were measured on plants from different stages, unless a particular stage is by definition required to measure a trait such as for seed mass (i.e. seed stage), or Hmax (i.e. adult stage). Finally, we weighted our data by number of species rather than the sample size for all statistical analyses (see Statistical analyses section).

@@

We acknowledge that it would be better to address plant size as a continous variable, but the limited number of correlations available in the literature along plant size steered us to focus on plant stage as an approximation. 
The authors of the 112 articles reported in our meta-analyses, categorized plants into four differents stages (i.e. seedling, juvenile, sapling, adult). They also reported plant size informations in diameter, height or age; generally the authors give the size of the smallest and the biggest plants species used in their analyses. While the stage categories have been established according to the expertise and the appreciation of the authors, by compiling data across multiple studies and across different growth form (tree, woody, etc), there is a risk of assigning plants of similar size to different categories. Thus, we tested how consistent were the range values in size/age attributed to each stage across the 112 studies, and across the different growth form (tree, woody and across growth form, Fig. A1). Suprisingly, the threshold in size/age used by the authors to defined a plant stage are quite steady across the 112 studies (Fig. A1). In particular, there was a clear break between seedlings and saplings in both age (at 1 yr) and height (at 0.5 m); there was also a clear break between saplings and adults in diameter (at 10 cm). The distinctions between sapling-adult in age (around 5 yrs) and height (around 2 m), and between seedling-sapling in diameter (around 1cm) were less pronounced. We used simple allometric relationships from the Biomass And Allometry Database (BAAD) \citep{Falster:2015} to ensure that these size/stage cutoffs match across the different size measurements (i.e. diameter, height and age, Fig. A2). For instance, we showed according to BAAD that a sapling with a diameter < 10 cm would have a height < 1 m. Thus, according to the size/stage threshold established across 112 studies, the plant will be assigned to the sapling stage whatever we focused on its diameter or height (see Fig. A2 c).

@@

Results using the categories established by the authors are similar than the one using the standardized categories and are available in Fig. A8. In the main text, we report the results using the standardized categories. In the rest of the mansucript, we preferred the term "juvenile" to "seedling", because for a portion of the scientific community "seedling" also refers to a particular experimental condition (a greenhouse experiment).

@@

We tested stage effect on traits for which > 15 correlation coefficients across at least two stages were recorded. We used a mixed-effect model of meta-analysis, in which it is assumed that studies within a stage share a common mean effect but that there is also random variation among studies in a stage, in addition to sampling variation. This random effect allows us to account for the heterogeneity among studies (e.g.~experimental type, vegetation type, error, etc.). For each trait, we established if stage could explain variance in the coefficient correlation by comparing models with the stage variable to the null model using log-likelihood ratios \citep{Zuur:2009cfa}. If the log-likelihood test returned a significant p-value, the stage model fitted the data better than the null model. Estimates of the effect size for each stage were considered significant if the 95\% confidence intervals did not overlap with zero. In the case of multiple comparisons within a study, we used the method proposed by \citet{Borenstein:2009um}: we calculated the mean correlation coefficient by trait and by study, and re-calculated the effect size. Finally, we checked the dataset for publication bias with funnel plots (\citealt{Koricheva:2013tz}, Fig. A9). A publication bias toward significant results creates an asymmetric funnel with small studies with non-significant effects missing from the mouth of the funnel on the side opposite to the true effect. In addition, to avoid publication bias due to selective reporting of significant results within published articles (e.g.~particular case of a 'failure to publish bias' see \citealt{Jennions:2013ta}), we asked the authors for additional information when they reported a non-significant correlation between growth and trait without reporting an r value. Finally, we explored the temporal change im effect size, by testing the effect of year of publication within stage for each trait. In Table 2 in Appendix, we reported how and where (i.e in the main text and/or figures) we addressed the 14 methodological quality criteria for meta-analysis listed by Koricheva and Gurevitch (2014).

@@

We were pleasantly surprised to find that correlations had been reported for at least two different stages across 18 traits out of the 20 targeted (Fig. \ref{Fig1}). We had thought it likely that for some traits correlations would be reported only for one size class, or only mixed across different sizes. However these limitations applied only for LAR and NARmass, where studies appeared confined to juvenile stages.

We found 112 studies published between between 1983 and 2014 that met our requirements. Climate zones spanned from temperate regions to  tropics (map in Fig. A3). Different experiment types (laboratory, field, plantation or greenhouse), growth measurements and growth form (tree, woody or across growth form) were included. Given this variety, it was not surprising to find wide ranges of r reported for all traits. For instance, r varied from -0.98 to 0.99 for SLA, and from -0.74 to 0.74 for WD (Fig. A4). The growth form, the experiment types and the growth measurement were not markedly best predictors of the data (results shown in the Fig. A8). Number of species used for each correlation ranged from 2 to 300 species with average 31. As expected the correlation coefficient varied less widely as sample size increased (funnel plots Fig. A9). In the analyses described in Figs 2, A5 and A6 coefficients were weighted according to sample size, making the small-sample extremes less influential. Funnel plots exhibited a typical funnel shape for all five functional traits analyzed, showing no evidence for publication bias (Fig. A9).

>>>>

\item
Growth rates have often been seen as central elements of plant ecological strategies, and as linked to species traits. Yet, the literature is inconsistent about empirical correlations between functional traits and plant growth rate (GR), casting doubt on the capacity of traits to predict growth.

@@

\section*{Introduction}\label{introduction}
The phrase "plant ecological strategy" refers to how species face the challenge they all share:  to acquire sufficient water, nutrients and light for individual growth and to live long enough to replace themselves with offspring. An idea that species strategy differences might involve fast vs slow growth goes back at least to Grime and Hunt \citeyearpar{Grime:1975gr}, who measured a wide spread of relative growth rates for seedlings during early exponential growth under favourable conditions. Seedling potential relative growth rate (RGR, the change in a mass/height or diamater per time interval relative to initial size) has continued to be used as a major strategy indicator within the CSR (Competitor - Stress tolerator - Ruderal) scheme (e.g. \citealt{grime1979plant, Grime:1997wm}). Meanwhile partitioning of seedling RGR into its components (see equation X in Appendix 1) showed that specific leaf area (SLA = 1/leaf mass per area LMA) was typically the strongest source of variation between species \citep{Poorter:1989tx}. From the 1990s a new style of "trait ecology" arose whereby measurable traits including SLA were used directly as strategy axes \citep{Westoby:2002ft}. An approach via measurable traits made broad comparisons possible across continents and latitudes and thousands of species. Traits have been seen as constructive approaches to understanding plant growth strategies \citep{Grime:1977kc,Chapin:1980gz}, community composition \citep{Lavorel:2002ff,Shipley:2006ie}, global vegetation dynamics \citep{Scheiter:2013ed} and ecosystem processes \citep{Lavorel:2002ff}.

Currently two main spectra of variation are widely cited as underpinning differences between species in vegetative growth rates. The 'leaf economic spectrum' \citep{Wright:2004jb} and the 'wood economic spectrum' \citep{Chave:2009iy} reflect tissue construction costs (for specific leaf area and for wood volume respectively) trading off with tissue turnover rates or mortality risks. It has been widely expected that species with low tissue construction cost will typically have fast growth rates at least in favourable physical environments (e.g. \citealt{MullerLandau:2004dc,Wright:2004jb,Poorter:2008iu,Chave:2009iy,Larjavaara:2010bn,Iida:2012jb,Paine:2015df}). Yet a growing number of empirical studies indicate the correlation between traits and growth is not as consistent as that. Consider SLA as an illustration. Species with high SLA can deploy more canopy area for a given dry mass investment in leaves \citep{Poorter:1999wd, Reich:1992wm}. In addition, SLA is correlated with high photosynthetic rates and high nitrogen (N) or phosphorus concentrations in the leaf \citep{Wright:2004jb}. For seedling relative growth rates under favourable conditions, high SLA has repeatedly been found to be a strong predictor regardless of vegetation type or growth form \citep{Lambers:1992bj,Reich:1992wm,Grime:1997wm,Poorter:1999wd,Wright:1999ds}. Yet for adult plants, this correlation appears weaker or absent \citep{coomes_comparison_1998,Poorter:2008iu,Aiba:2009ft,Easdale:2009gv,Wright:2010tp}. The growing discrepancy between theoretical expectation and empirical results has led some to question whether traits are even useful for understanding growth \citep{Wright:2010tp, Paine:2015df}.

Meanwhile the possibility has been raised that trait-growth correlations may change systematically with size Inquist et al () \citep{Falster:2011ii, Ruger:2012jv, Iida:2014ep, Iida:2014hq}. Recent theory suggests that the absence of a correlation between SLA and growth at larger sizes may be expected, when potential influences of traits on plant function are considered \citep{Falster:2011ii, falster:2013}. At larger sizes the low-cost deployment benefits of high SLA may be offset or even outweighed by the high costs of rapid leaf turnover (see Text Box 1\ref{sec:growth}).

@@

The Web of Science database (Thomson Reuters) was searched between March and June 2014 using a range of Boolean search terms in English. Search terms were as follows: (''NAME'' AND (''growth rate'' OR ''relative growth rate'')) AND (''tree*'' OR ''woody*''), ''NAME'' was one of trait names listed in TableA1. According to this snapshot, 62\% of the literature for the 20 traits (or 82\% for the 5 traits for which we have clear predictions) is captured by searching only on SLA and WD.

@@

Google Scholar database was searched using the range of Boolean search terms for SLA and WD. Search terms were as follows: ((''SLA'' OR ''LMA'' OR ''specific leaf area'' OR ''leaf mass area'' OR ''wood density'' OR ''WD'' OR ''wood specific gravity'' OR ''stem density'' OR ''stem specific density'') AND (''growth rate'' OR ''relative growth rate'')) AND (tree* OR woody*). The keyword search in Google Scholar returns approx. 18900 papers. For comparison, with the same keyword search Web of Science returns 541 papers. Thus, in Google Scholar we displayed papers by relevance and examined the first 750 items. 
Each study had to meet some basic criteria to be included in our analysis. We restricted our meta-analysis to documents reporting: i) the relationship between growth and trait as a correlation coefficient, ii) the sample size and the number of species used to perform the correlation, and iii) the size/stage of plants. Where available we also recorded the geographic coordinates, number of sites, experiment type (greenhouse, field experiment, forest, shade-house etc.), growth measure (relative or absolute growth rate measured using plant diameter, height or mass), vegetation type, life form (tree, woody or mixed with herbaceous species etc.), biological scale (interspecific, intraspecific or intragenus), and experimental treatments (e.g.~fertilization, shade, salt stress). Ultimately data was included from 112 different studies (starred references in bibliography). The articles laid aside because they did not meet our criteria are listed in Appendix \ref{app:supp_info_files} in Supplementary material.

The number of species used to perform the correlation, the experimental conditions, and the concordance between stage used to measure trait and growth are clear criteria for assessing the quality of the data for addressing our hypotheses. Therefore, we focused on correlation across species, and we stratified the data set into two subsets: one including exclusively studies under unstressed conditions ('ideal' dataset), and the other set including studies under all conditions ('complete' dataset). We excluded from the 'ideal' dataset studies where growth and traits were measured on plants from different stages, unless a particular stage is by definition required to measure a trait such as for seed mass (i.e. seed stage), or Hmax (i.e. adult stage). We weighted our data by number of species rather than the sample size for all statistical analyses (see Statistical analyses section). Finally, to avoid publication bias due to selective reporting of significant results within published articles (e.g.~particular case of a 'failure to publish bias' see \citealt{Jennions:2013ta}), we asked the authors for additional information when they reported a non-significant correlation between growth and trait without reporting an r value.

@@

We acknowledge that it would be better to address plant size as a continous variable, but the limited number of correlations available in the literature along plant size constrains us to approximate plant size by plant stage. The authors of the 112 articles reported in our meta-analyses, categorized plants into four differents stages (i.e. seedling, juvenile, sapling, adult). They also reported plant size informations in diameter, height or age; generally the authors give the size of the smallest and the biggest plants species used in their analyses. While the stage categories have been established according to the expertise and the appreciation of the authors, by compiling data across multiple studies and across different growth form (tree, woody, etc), there is a risk of assigning plants of similar size to different categories. Thus, we tested how consistent were the range values in size/age attributed to each stage across the 112 studies, and across the different growth form (tree, woody and across growth form, Fig. A1). Suprisingly, the threshold in size/age used by the authors to defined a plant stage are quite steady across the 112 studies (Fig. A1). In particular, there was a clear break between seedlings and saplings in both age (at 1 yr) and height (at 0.5 m); there was also a clear break between saplings and adults in diameter (at 10 cm). The distinctions between sapling-adult in age (around 5 yrs) and height (around 2 m), and between seedling-sapling in diameter (around 1cm) were less pronounced. We used simple allometric relationships from the Biomass And Allometry Database (BAAD) \citep{Falster:2015} to ensure that these size/stage cutoffs match across the different size measurements (i.e. diameter, height and age, Fig. A2). For instance, we showed according to BAAD that a sapling with a diameter < 10 cm would have a height < 1 m. Thus, according to the size/stage threshold established across 112 studies, the plant will be assigned to the sapling stage whatever we focused on its diameter or height (see Fig. A2 c).

@@

Results using the categories established by the authors are similar than the one using the standardized categories and are available in Fig. A8 model.1. In the main text, we report the results using the standardized categories. In the rest of the mansucript, we preferred the term "juvenile" to "seedling", because for a portion of the scientific community "seedling" also refers to a particular experimental condition (a greenhouse experiment).

@@

We tested stage effect on traits for which > 15 correlation coefficients across at least two stages were recorded. We used a mixed-effect model of meta-analysis, in which it is assumed that studies within a stage share a common mean effect but that there is also random variation among studies in a stage, in addition to sampling variation. This random effect allows us to account for the heterogeneity among studies (e.g.~experimental type, vegetation type, error, etc.). For each trait, we established if stage could explain variance in the coefficient correlation by comparing models with the stage variable to the null model using log-likelihood ratios \citep{Zuur:2009cfa}. If the log-likelihood test returned a significant p-value, the stage model fitted the data better than the null model. Estimates of the effect size for each stage were considered significant if the 95\% confidence intervals did not overlap with zero. In the case of multiple comparisons within a study, we used the method proposed by \citet{Borenstein:2009um}: we calculated the mean correlation coefficient by trait and by study, and re-calculated the effect size.

Following recent call for the adoption of a checklist of quality criteria for meta-analysis in ecology (Koricheva and Gurevitch, 2014), we conducted several analyses to assess the reliability of our meta-analysis. Table A2 in Appendix X reported how and where (i.e in the text and/or figures) we addressed the 14 methodological quality criteria listed by Koricheva and Gurevitch (2014). In particular, we checked the dataset for publication bias with funnel plots (\citealt{Koricheva:2013tz}) and the 'trim and fill' method (Appendix), we explored the temporal changes in effect size (Appendix), and we estimated the extend of heterogeneity using the statistic I2 (Appendix X).

@@

We were pleasantly surprised to find that correlations had been reported for at least two different stages across 18 traits out of the 20 targeted (Fig. \ref{Fig1}). We had thought it likely that for some traits correlations would be reported only for one size class, or only mixed across different sizes. These limitations applied only for LAR and NARmass, where studies appeared confined to juvenile stages. However, among the 20 traits studied we had enough data to test the effect of stage on the trauts value only for the five for which we had predictions.

We found 112 studies published between between 1983 and 2014 that met our requirements. Climate zones spanned from temperate regions to  tropics (map in Fig. A3). Different experiment types (laboratory, field, plantation or greenhouse), growth measurements and growth form (tree, woody or across growth form) were included. Given this variety, it was not surprising to find wide ranges of r reported for all traits. For instance, r varied from -0.98 to 0.99 for SLA, and from -0.74 to 0.74 for WD (Fig. A4). The growth form, the experiment types and the growth measurement were not markedly best predictors of the data (results shown in the Fig. A8). Number of species used for each correlation ranged from 2 to 300 species with average 31.

Finally, we met all the quality criteria required for meta-analysis in ecology. We found no evidence of publication bias, the extend of the heterogenity appares to be low to moderate in our meta-analysis, and there is no temporal change in effect size calculated within a stage.
